The curious world of constructive mathematics

### 2022-09-03 algar-lecture-notes.tex
Beginnings on extracting content from classical proofs

\usepackage[natbib=true,style=numeric,maxnames=10]{biblatex}
\usepackage[babel]{csquotes}
\bibliography{bibliography}

@@

%\begin{bem}Die Formulierung mit dem generischen \emph{Wir} in Tafel~\ref{bhk}
ist etwas irreführend. Wie auch in klassischer Mathematik hängen
intuitionistische Urteile nicht von \emph{uns} oder anderen Mathematikern ab.
Aussagen, die bisher noch nicht konstruktiv bewiesen wurden, können
durchaus einen (noch unbekannten) Beleg besitzen. Eine genauere Diskussion
findet sich etwa in~\cite[Seite~42f.]{kohlenbach:applprooftheory}. Die
Brouwer--Heyting--Kolmogorov-Interpretation kann im Rahmen der
\emph{Realisierbarkeitstheorie} formalisiert werden~\cite{bauer:realizability};
ohne eine solche Formalisierung gibt es gewisse weitere
Probleme~\cite{artemov:bhk,sanz:piecha:critical-bhk,dalen:bhk}.
%\end{bem}

@@

{\noindent\scriptsize\emph{Remark.} The set~$\NN_\infty$ is the \emph{one-point
compactification} of~$\NN$. A sensible definition in constructive mathematics
is as the set of decreasing binary sequences~$(x_0,x_1,x_2,\ldots)$. The
naturals embed into~$\NN_\infty$ by mapping~$n$ to the sequence~$1^n 0^\omega =
(1,\ldots,1,0,\ldots,0)$, and an element not in the image of this embedding
is~$\infty \defeq 1^\omega = (1,1,\ldots)$. Assuming the law of excluded
middle (or already weaker principles), every element of~$\NN_\infty$ is of one
of these two forms. Martín Escardó has worked extensively on unexpected
instances of the principle of omniscience for searchable sets
like~$\NN_\infty$~\cite{escardo:omniscient1,escardo:omniscient2}.\par}
\end{enumerate}
\end{exercise}

@@

{\noindent\scriptsize\emph{Note.} Strictly speaking, this exercise presupposes
familiarity with an intuitionistic account of the basics of undergraduate real
analysis. Without it, one cannot really be expected to precisely think about
these matters. One such account (though assuming the axiom of dependent choice)
is~\cite{bridges}. However, this exercise is insightful even when carried out
slightly informally. Keep in mind that, to show that a real number is positive,
constructively it is not enough to merely verify that it cannot be zero or
negative. A safe way to verify that a real number~$a$ is positive is to exhibit
a rational number~$b$ such that~$a \geq b > 0$.\par}
\end{exercise}

@@

@@

omniscient1
https://www.cs.bham.ac.uk/~mhe/.talks/dagstuhl2011/omniscient.pdf

omniscient2
https://www.ioc.ee/~tarmo/tsem16/escardo2605-slides.pdf

>>>>

\usepackage[natbib=true,style=numeric,maxnames=10]{biblatex}
\usepackage[babel]{csquotes}
\bibliography{algar-lecture-notes}

@@

%\begin{bem}Die Formulierung mit dem generischen \emph{Wir} in Tafel~\ref{bhk}
ist etwas irreführend. Wie auch in klassischer Mathematik hängen
intuitionistische Urteile nicht von \emph{uns} oder anderen Mathematikern ab.
Aussagen, die bisher noch nicht konstruktiv bewiesen wurden, können
durchaus einen (noch unbekannten) Beleg besitzen. Eine genauere Diskussion
findet sich etwa in~\cite[Seite~42f.]{kohlenbach:applprooftheory}. Die
Brouwer--Heyting--Kolmogorov-Interpretation kann im Rahmen der
\emph{Realisierbarkeitstheorie} formalisiert werden~\cite{bauer:realizability};
ohne eine solche Formalisierung gibt es gewisse weitere
Probleme~\cite{artemov:bhk,sanz-piecha:critical-bhk,dalen:bhk}.
%\end{bem}

@@

{\noindent\scriptsize\emph{Remark.} The set~$\NN_\infty$ is the \emph{one-point
compactification} of~$\NN$. A sensible definition in constructive mathematics
is as the set of decreasing binary sequences~$(x_0,x_1,x_2,\ldots)$. The
naturals embed into~$\NN_\infty$ by mapping~$n$ to the sequence~$1^n 0^\omega =
(1,\ldots,1,0,\ldots,0)$, and an element not in the image of this embedding
is~$\infty \defeq 1^\omega = (1,1,\ldots)$. Assuming the law of excluded
middle (or already weaker principles), every element of~$\NN_\infty$ is of one
of these two forms. Martín Escardó has worked extensively on unexpected
instances of the principle of omniscience for searchable sets
like~$\NN_\infty$~\cite{escardo:omniscience1,escardo:omniscience2,escardo:omniscience}.\par}
\end{enumerate}
\end{exercise}

@@

{\noindent\scriptsize\emph{Note.} Strictly speaking, this exercise presupposes
familiarity with an intuitionistic account of the basics of undergraduate real
analysis. Without it, one cannot really be expected to precisely think about
these matters. One such account (though assuming the axiom of dependent choice)
is~\cite{bishop-bridges:bible}. However, this exercise is insightful even when carried out
slightly informally. Keep in mind that, to show that a real number is positive,
constructively it is not enough to merely verify that it cannot be zero or
negative. A safe way to verify that a real number~$a$ is positive is to exhibit
a rational number~$b$ such that~$a \geq b > 0$.\par}
\end{exercise}

@@

\begin{thm}[Dickson's lemma]\label{thm:dickson}
Let~$k \in \NN$. Let~$f : \NN \to \NN^k$ be an arbitrary map. Then there are
indices~$i < j$ such that~$f(i) \leq f(j)$ (componentwise).
\end{thm}

\begin{proof}[Proof (classical)]The case~$k = 0$ is trivial and we omit the cases~$k
\geq 2$, hence let~$k = 1$. In this case, the map~$f$ attains some minimal
value. Set~$i$ to be (one of the) positions where this minimal value is
attained. Set~$j \defeq i+1$. Then, trivially, $f(i) \leq f(j)$.
\end{proof}

\begin{thm}\label{thm:surjective-matrix}
Let~$M$ be a surjective matrix with more rows than columns over a commutative
ring~$A$ with unit. Then~$1 = 0$ in~$A$.
\end{thm}

\begin{proof}[Proof (classical)]Assume not. Then there is a maximal
ideal~$\mmm \subseteq A$. The matrix~$M$ remains surjective when considered over
the quotient ring~$A/\mmm$, and by maximality this quotient ring is a field.
Hence we have a contradiction to basic linear algebra, namely to the basic fact
that matrices over fields are not surjective if they have more rows than
columns.\end{proof}

What is the meaning of these non-effective proofs? Theorem~\ref{thm:dickson}
claims the existence of a finite object with a decidable property (a
pair~$(i,j)$ such that~$f(i) \leq f(j)$), but the given proof employs
transfinite methods and gives no indication how we could compute or otherwise
find this object. Instead, the classical proof asks us to grasp the infinitude
of all values of~$f$ so we can determine the minimum. The issue is even more
pronounced with Theorem~\ref{thm:surjective-matrix}, since the existence of
maximal ideals in nontrivial rings requires the axiom of
choice~\cite{scott:prime-ideals,hodges:krull,banaschewski:krull,erne:krull,howard-rubin:ac}.
To add one more conundrum: Assume that we have a classical proof, using the law
of excluded middle and the axiom of choice, that some given Turing machine
terminates. Can we then constructively accept that the machine will halt? Do we
have an upper bound for the number of computational steps the machine carries
out before halting?

Astoundingly, it is almost always the case that from classical proofs useful
constructive content can be extracted. In fact, due to general metatheorems,
in many cases there are even explicit mechanical procedures for extracting this
hidden content, while other cases require more creativity for determining
suitable constructive reformulations. For instance:

% \begin{document}

\begin{enumerate}
\item \emph{Eliminating the axiom of choice by the~$L$-translation.} Can the
axiom of choice ever help in proving arithmetical statements, those first-order
statements in which all quantifiers range over the natural numbers? Well, it
might. But a result of Gödel states that~\textsc{zfc} (Zermelo--Fraenkel set
theory with the axiom of choice) is conservative over~\textsc{zf} (ZF set
theory without it) for such statements -- hence all appeals to this axiom can
be mechanically eliminated from a given proof. This is true even if the proof
transcends the arithmetical realm and includes statements which are not
arithmetical, only the asserted claim is required to be arithmetical. Hence we
are free to use the axiom of choice, tranquil in knowing that we could always
reformulate our proofs without it.\footnote{Zermelo--Fraenkel set theory with
the axiom of choice is the go-to foundation of mathematics often cited as
supporting ``almost all'' of current mathematics, one important exception being
some (definitions and) results in category theory dealing with large
structures~\cite{shulman:set-theory,feferman:set-foundations}. A fundamental
result due to Gödel is that the axiom of choice ``holds in~$L$'', the
\emph{constructible universe}, even if it might not hold in~$V$, the true
universe of all sets. More precisely, if~\textsc{zfc} shows some
statement~$\varphi$, then~\textsc{zf} shows its~$L$-relativized
version~$\varphi^L$, where all quantifiers have been restricted to range
over~$L$ instead of~$V$. The conservation result follows because the natural
numbers ``are absolute between~$V$
and~$L$''~\cite{goedel:ac-gch,schoenfield:predicativity}.}

\item \emph{Eliminating the law of excluded middle by the double-negation
translation and its variants.} At the price of
slightly modifying the asserted claim, the law of excluded middle
can always be mechanically eliminated from a given proof. This
elimination procedure is facilitated by the \emph{double-negation translation}
reviewed below. In some cases, a refined translation even allows to preserve the asserted
claim exactly; this technique is variously known as \emph{Friedman's trick},
\emph{nontrivial exit continuation} or (the baby version of) \emph{Barr's
theorem}. To cite a concrete instance of this phenomenon, classical~\textsc{zf}
set theory is conservative over its intuitionistic cousin~\textsc{izf}
for~$\Pi^0_2$-statements (statements of the form~$\forall\ldots\forall\_
\exists\ldots\exists\_ \%$, where all quantifiers in the final~``$\%$''
are bounded).

\item \emph{Embracing generic models.} A useful companion to both of the
aforementioned techniques is to switch from referencing all models of a certain
kind to referencing only the \emph{generic model}. For instance, Krull's lemma
stating that a ring element is already nilpotent if it is contained in all
prime ideals requires the Boolean Prime Ideal Theorem, a slightly weaker
version of the axiom of choice but still ineffectiv and unconstructive.
However, Krull's lemma is valid in the form that a ring element is nilpotent if
it is contained in the \emph{generic prime ideal}. This particular example has
received lots of attention (see the references
in~\cite{blechschmidt-schuster:constructive-maximal-ideals}) and the general
technique will be the object of the fourth lecture.
\end{enumerate}

@@
